No one-size-fits-all solution to clean GBIF

### 2020-08-04 manuscript_peerj.tex
2nd resubmission

% Pandoc Header
\usepackage[export]{adjustbox}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses (for instance because they are overly imprecise, Anderson et al. 2016). Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times--- often without centralized curation and only rudimentary meta-data. For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al. 2015; Yesson et al. 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity (Zizka et al. 2019).

@@

This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group. Hence, we include groups based on the expertise of the participants rather than following an arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree of documented expertise and number of years working on each group. We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p,2020a,b). We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90\(^\circ\)S - 33\(^\circ\) N and 35\(^\circ\) W - 120\(^\circ\) W reflecting the Neotropics (Morrone 2014), our main area of expertise. The natural distributions of all included taxa are confined to the Neotropics except for Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia}, for which we only obtained the Neotropical occurrences. We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist (e.g., \url{https://bien.nceas.ucsb.edu/bien/}, www.fishbase.de or Guedes et al. 2018). The variety of existing databases is also the reason why we downloaded data without accounting for GBIF-internal quality testing (``Geospatial issues''), since quality control mechanisms might differ among databases and sometimes quality control might not be carried out (for instance in \emph{de novo} databases compiled by groups of researchers).

@@

We followed the cleaning pipeline outlined by Zizka et al. (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although meta-data were often missing, Peterson et al. 2018), removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka et al. 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all filters we used and their classification into ``erroneous'' and ``unfit''.

@@

Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2). However, in some cases filters failed to remove major issues, often due to incomplete metadata. For instance, for Diogenidae we found at least two records of an extinct species (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''. Furthermore, for Entomobryidae we found that for 1,996 records the meta-data on taxonomic rank was ``UNRANKED'' despite all of them being identified to species level, leading to a high fraction of records removed by the ``Identification level'' filter. Additionally automated filters might be overly strict or unsuitable for certain taxa. For instance, in Entomobryidae, 2,004 samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.

The filters we included in this study address a set of important but relatively easy to identify problems. In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}, \url{https://data-blog.gbif.org/post/country-centroids/}). While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why). Such a system would not need to change the original data and could include multiple levels to account for differing opinions.

As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families). Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.

@@

Our results suggest that between one to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses. While the majority of these records might not be erroneous \emph{per se}, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.

While our results suggest that large-scale species richness patterns and automated conservation assessments are largely resilient to the effects of problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses. Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest. The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide data aggregators such as GBIF with feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.

@@

\caption{\label{tab:tabletaxa}The study groups and their taxonomy. This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders. The outlines illustrate the broad taxonomic group. Pictures from www.phylopic.org. Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T. Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \textit{Gaylussacia}, \textit{Lepismium}, \textit{Pilosocereus}: T. Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \textit{Iridaceae} adapted from Goldblatt and Manning (2008); \textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.}
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}}
 & Taxon & Taxon rank & Realm & Common name & 'Phylum' & Family\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/diogenidae.png} & Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Diogenidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/arhynchobatidae.png} & Arhynchobatidae & Family & Marine & Skates & Chordata & Arhynchobatidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
\includegraphics[valign=T, scale=0.18, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/thozetella.png} & \em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/conchocarpus.png} & \em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Rutaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/gaylussacia.png} & \em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
\includegraphics[valign=T, scale=0.075, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/oocephalus.png} & \em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/pilosocereus.png} & \em{Pilosocereus} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/prosthechea.png} & \em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Orchidaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\

@@

\caption{\label{tab:tabletests}The automated filters used in this study.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{9cm}}
\toprule
Test & Type & Basis & Rationale\\
\midrule
\rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g. because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
\rowcolor{gray!6}  Sea & Error & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
Zeros & Error & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
\rowcolor{gray!6}  Capitals & Unfit & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
\addlinespace
Duplicates & Unfit & Gazetteer-based & Duplicated records may add unnecessary computational burden, in particular for large scale biodiversity analyses and distribution modelling for many species\\
\rowcolor{gray!6}  Political centroids & Unfit & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. 'fossil' or 'unknown'\\
Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description. This is more problematic for older records, since names or borders of places may change\\
\addlinespace
\rowcolor{gray!6}  Coordinate precision & Unfit & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
Identification level & Unfit & Meta-data & Records may be unsuitable if they are not identified to species level.\\
\rowcolor{gray!6}  Individual count & Unfit & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is too high, as this is often related to records from barcoding or indicative of data entry problems.\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups_reduced} \caption{Examples of taxa included in this study. \textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira.}\label{fig:species}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics.}\label{fig:total}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./output/figure_number_of_records_split} \caption{Geographic location of the occurrence records flagged by the automated tests applied in this study. Only filters that flagged at least 0.1\% of records in any taxon are shown.}\label{fig:split}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./figures_tables/figure_number_species_richness_difference} \caption{Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa. Photo credits for C) by Tiago Andrade Borges Santos, otherwise as in Figure 1.}\label{fig:speciesrichness}
\end{figure}

>>>>

% Pandoc Header
\usepackage[export]{adjustbox}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.5 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses (for instance because they are overly imprecise, Anderson et al. 2016). Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times---often without centralized curation and only rudimentary meta-data. For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al. 2015; Yesson et al. 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity (Zizka et al. 2019).

@@

This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics (Fig. \ref{fig:species}). Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group. Hence, we include groups based on the expertise of the participants rather than following an arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree of documented expertise and number of years working on each group. We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p,2020a,b). We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90\(^\circ\)S - 33\(^\circ\) N and 35\(^\circ\) W - 120\(^\circ\) W reflecting the Neotropics (Morrone 2014), our main area of expertise. The natural distributions of all included taxa are confined to the Neotropics except for Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia}, for which we only obtained the Neotropical occurrences. We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist (e.g., \url{https://bien.nceas.ucsb.edu/bien/}, www.fishbase.de or Guedes et al. 2018). GBIF provides information on the internal consistency of records, among others including information on decimal rounding of coordinates, geographic projection and date validity and geospatial issues (including the zero coordinates test used in this study). Since we specifically aimed to test the effect of user-level filtering we included records flagged with issues by GBIF (this was also the default option). Geospatial issues flagged by GBIF only concerned 0.4\% of the records used in this study and including them had the added benefit to make our results directly comparable to other databases, which may use different internal consistency checks or none at all.

@@

We followed the cleaning pipeline outlined by Zizka et al. (2019) and first filtered the data as downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although meta-data were often missing, Peterson et al. 2018), removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka et al. 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all filters we used and their classification into ``erroneous'' and ``unfit''.

@@

Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2). However, in some cases filters failed to remove major issues, often due to incomplete meta-data. For instance, for Diogenidae we found at least two records of an species known only from Eocene fossils (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''. Furthermore, for Entomobryidae we found that for 1,996 records the meta-data on taxonomic rank was ``UNRANKED'' despite all of them being identified to species level, leading to a high fraction of records removed by the ``Identification level'' filter. Additionally automated filters might be overly strict or unsuitable for certain taxa. For instance, in Entomobryidae, 2,004 samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.

The filters we included in this study address a set of important but relatively easy to identify problems. In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}). While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why). Such a system would not need to change the original data and could include multiple levels to account for differing opinions.

As next steps for automated filtering, tests for intrinsic consistency and support by external data (if available) can help to detect additional problematic records. For instance, testing if records' coordinates fall within the state or province of collection noted for a record (intrinsic) or agree with external species distribution information, for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families; extrinsic) can further corroborate the accuracy of a record's geographic referencing. If such tests are included it is essential to account for the sampling year, in particular for older records, since the names of provinces may change and the ranges of species may shift. Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.

@@

Our results suggest that between one quarter to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses. While the majority of these records might not be erroneous \emph{per se}, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.

While our results suggest that large-scale species richness patterns and automated conservation assessments are largely resilient to the effects of problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses. Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest. The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to visibly associate user-feedback with individual records on aggregator platforms such as GBIF so that it can contribute to the overall accuracy and precision of public biodiversity databases.

@@

\caption{\label{tab:tabletaxa}The study groups and their taxonomy. This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders. The outlines illustrate the broad taxonomic group. Pictures from www.phylopic.org if available under a Public Domain Dedication license, otherwise generated from photographs available in the public domain.}
\centering
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}}
\toprule
 & Taxon & Taxon rank & Realm & Common name & 'Phylum' & Family\\
\midrule
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/diogenidae.png} & Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Diogenidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/arhynchobatidae.png} & Arhynchobatidae & Family & Marine & Skates & Chordata & Arhynchobatidae\\
\addlinespace
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
\includegraphics[valign=T, scale=0.18, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
\hline
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/thozetella.png} & \em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriaceae\\
\hline
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/conchocarpus.png} & \em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Rutaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/gaylussacia.png} & \em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericaceae\\
\addlinespace
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
\includegraphics[valign=T, scale=0.075, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/oocephalus.png} & \em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/pilosocereus.png} & \em{Pilosocereus} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\addlinespace
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/prosthechea.png} & \em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Orchidaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:tabletests}The automated filters used in this study.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{9cm}}
\toprule
Test & Type & Basis & Rationale\\
\midrule
\rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g. because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture, which are not clearly labeled as such\\
Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
\rowcolor{gray!6}  Sea & Error & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
Zeros & Error & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
\rowcolor{gray!6}  Capitals & Unfit & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
\addlinespace
Duplicates & Unfit & Gazetteer-based & Duplicated records may add unnecessary computational burden, in particular for large scale biodiversity analyses and distribution modelling for many species\\
\rowcolor{gray!6}  Political centroids & Unfit & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. 'fossil' or 'unknown'\\
Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description. This is more problematic for older records, since names or borders of places may change\\
\addlinespace
\rowcolor{gray!6}  Coordinate precision & Unfit & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
Identification level & Unfit & Meta-data & Records may be unsuitable if they are not identified to species level.\\
\rowcolor{gray!6}  Individual count & Unfit & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is too high, as this is often related to records from barcoding or indicative of data entry problems.\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{./figures_tables/Fig1_study_groups} \caption{Examples of taxa included in this study. \textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Calisto Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur de Souza Soares, J) Renata C Santos-Costa, K) Tiago Vieira.}\label{fig:species}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./figures_tables/Fig2_number_of_records} \caption{The absolute number of records flagged as erroneous or unfit by automated geographic filters in dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics (Behrmann projection).}\label{fig:total}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./figures_tables/Fig3_number_of_records_split} \caption{Geographic location of the occurrence records flagged by the automated tests applied in this study. Only filters that flagged at least 0.1\% of records in any taxon are shown.}\label{fig:split}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./figures_tables/Fig4_number_species_richness_difference} \caption{Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa. Total species number in the raw data sets: Dipsadidae: 637, \textit{Harengula}: 4, \textit{Thozetella}: 9, \textit{Tillandsia} 464.  Photo credits for C) by Tiago Andrade Borges Santos, otherwise as in Figure 1.}\label{fig:speciesrichness}
\end{figure}
